---
title: "Independent Project Model Specification"
author: "Wynne Moss"
date: "October 31, 2018"
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include = FALSE}
library(rstanarm)
library(ggplot2)
library(lme4)
library(rstan)
```

## The data
SUMMARY OF DATASET: I quantified parasite communities across 10 different pond sites in the East Bay of California; each site was visited 4-6 times within the 2017 summer. At each visit I collected 10-12 individuals from 2 amphibian species. Individuals were measured and parasite infection was quantified.

GROUPING VARIABLES: 

  * Site (10 ponds) - this is a random variable.
  
  * Visit (6 visits) - I want to treat this as a fixed effect, even though visits are nested (?) within sites (so should it go in as a predictor?)
  
  * Species (2 species) - Again, I want to treat this as a fixed effect, although species are nested(?) within site visit

PREDICTOR VARIABLES:

  * Species (2 species) - I think this is a better designation for species
  
  * Body size (snout-vent-length) - A continuous variable (fixed effect) at the individual level
  
  * Developmental stage - A continuous variable at the individual level...problematic since development is measured differently in newts vs. frogs
  
  * Sex - A factor variable at the individual level

RESPONSE VARIABLE:

  * The number of parasites found within an individual. For this analysis, just count number of Echinostoma parasites. Can be modeled as Poisson distributed with a negative binomial to account for aggregation. Infection status (1 or 0) could also be modeled as a binary response variable using logistic regression. 


## Questions
1) Does the impact of species and body size change over the course of the summer (interact with visit)?
2) How much of the variation in overall parasite load is explained by visit-level, species-level, site-level, or individual-level variation?

## Display the structure of the data
```{r}
dis <- read.csv("diss.data.2017.csv")
# colnames(dis)
str(dis)
```

## Model formulation
For now, I'll focus on question 1 with the response variable being `Echinostoma` (number of Echinostoma parasites found within that individual)

### Formulation 1
This is the model formulation for a random intercept model where visit is treated as a fixed effect (perhaps using julian date which is continuous)

$y_i$ ~ NB($\mu_{j[i]}$, $p$) 

$\mu_j[i]}$  is the rate (mean) and $p$ is the overdispersion parameter (since we can't let mean equal variance with overdispersed data).

I'm not sure if this equation is used, or whether $p$ is just modeled like any other parameter with maximum likelihood:
Var($y_i$) ~ $r_i$ + $r_i^2$/$p$  (Allows variance to increase with mean)

log($\mu_{j[i]}$) = $\alpha_{i[j]}$ +$\beta_1$SpeciesCode + $\beta_2$visit + $\beta_3}$(SpeciesCode x visit) + $\beta_4$SVL + $\beta_{5}$(SVL x Visit)

$\alpha_{i[j]}$ ~ Normal($\mu_\alpha$, $\sigma^2_\alpha$)
This is between pond variance. J's are ponds (10 total)

### Formulation 2
This is the model formulation for a random slope + random intercept model where visit is treated as a fixed effect

$y_i$ ~ NB($\mu_{i}$, $p$)

log($\mu_{i[j]}$) = $\alpha_{i[j]}$ +$\beta_1$SpeciesCode + $\beta_2$(SpeciesCode x Visit) + $\beta_3$SVL + $\beta_4$(SVL x Visit)  + $\beta_{i[j]}$visit

$\alpha_{i[j]}$ ~ Normal($\mu_\alpha$, $\sigma^2_\alpha$)

$\beta_{i[j]}$ ~ Normal($\mu_\beta$, ???)

Now, I'm not sure how these sigma's work since the random slopes and intercepts are correlated...
### Formulation 3
Instead of doing NB, can I have an observation-level random effect which accounts for dispersion? 

$y_{ij}$ ~ NB($\mu_{i}$, $p$)

log($\mu_{i[j]}$) = $\alpha_{i}$ +$\beta_1$SpeciesCode + $\beta_2$(SpeciesCode x Visit) + $\beta_3$SVL + $\beta_4$(SVL x Visit)  + $\beta_{i[j]}$visit

Individual level observations are modeled as a random effect, coming with some distribution centered around the mean for that individual...??? 

$\alpha_{i}$ ~ Normal($\alpha_{i[j]}$, $\sigma^2_\alpha$) What's the sigma here?

$\alpha_{i[j]}$ ~ Normal($\mu_j$, $\sigma^2_\alpha$) Each pond level alpha comes from the mean over all the ponds

$\beta_{i[j]}$ ~ Normal($\mu_\beta$, ???)

I confused myself...

### Formulation 4
I'm trying to incorporate Brett's feedback about using visit as another nesting (although I'm still really confused on nesting vs. crossing, since every site has the same 6 visits, the visit's don't seem unique to each site?)

It makes sense to me to just include both as random effects, but I think that the effect of visit DIFFERS by pond...

I think the model formula I'm looking for would be something like:
Echinostoma ~ (visit|SpeciesCode) + (visit|SVL) + (1|SiteCode)

But does this let me ask questions like: does the effect of species lessen over time?


## Stan formulation
In stan_glm, the model could be written as:
```{r, eval = FALSE}
stan.fit <- stan_glmer(Echinostoma ~ visit*SpeciesCode + visit*SVL + (1 + visit|SiteCode), data = dis,  family =neg_binomial_2(link="log"))
summary(stan.fit)
# launch_shinystan(stan.fit)
stan.samp <- sample(stan.fit)
samples <- extract(ppfit_bayes$stanfit)
```


Next steps: incorporate Brett's suggestions. Treat visit as a random effect (there are two levels of grouping), but perhaps have some other metric of visit (Julian date) as a fixed effect allowing it to interact with other variables to affect parasite count. Developmental stages could perhaps be included with an interaction with species, which would allow TATO to have some levels and PSRE to have other levels. e.g. I need to put them in the same column. Instead of using stan's negative binomial (which seems a bit problematic according to googling) I can include individual as a random effect.

- Will autocorrelation of fixed and random effects become an issue?
