---
title: "Simulation of parasite data for model checks"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rstan)
library(rstanarm)
library(tidyverse)
options(mc.cores = parallel::detectCores())
```

## Read in and clean up data
```{r}
ponddata <- read.csv("diss.data.2017.csv")
ponddata <- subset(ponddata, visit > 1) # take out first visit # fix this??
ponddata$visit <- ponddata$visit-1
ponddata <- ponddata[, c(2:6, 17,18)] # get just predictor variables
```

## SIMULATION 1: Site-level random effect
This model predicts parasite load using only a random effect for site.

Generate random effects for each of 10 sites from a normal distribution
```{r}
nsites <- length(levels(ponddata$SiteCode)) # number of levels of the random effect (10)
site.means.gamma <- NA
set.seed(12398) # makes sure the random draws are repeatable (?)

site.means.gamma <- rnorm(n = nsites, mean = 2, sd = .2) # 10 site-level means (log(mean parasite load))

# make a dataframe that we can merge with the predictor data frame
site.means.df <- data.frame(SiteCode = levels(ponddata$SiteCode), ExpMean = site.means.gamma)
# merge this with the existing predictors
ponddata2 <- merge(ponddata, site.means.df, by = "SiteCode")
```

Generate an observed parasite count from the expected site level mean. This is done using a negative binomial distribution.
```{r}
# for each row of the dataset, draw a random parasite count from the expected site level means
ponddata2$Echinostoma <- rnbinom(n = nrow(ponddata2), mu = exp(ponddata2$ExpMean), size = 1)
# using the negative binomial
# not really sure what size means here but it seems that smaller values make it more aggregated
```

Fit a model; does it recapture the random effects?
```{r}
stan.fit.sim1 <- stan_glmer(Echinostoma~ (1|SiteCode), data = ponddata2,  family =neg_binomial_2(link="log"))
summary(stan.fit.sim1)
rand.eff.sim1 <- coefficients(stan.fit.sim1)$SiteCode
rand.eff.sim1
site.means.gamma
```

Plot the estimates vs. real
```{r}
{plot(rand.eff.sim1[,1]~site.means.gamma, xlab = "Actual Random Effects", ylab = "Estimated Random Effects", xlim = c(1.5,2.5), ylim = c(1.5, 2.5))
abline(0,1)}
```


These look OK, now try again with diff set of random intercepts:
```{r}
set.seed(231)
site.means.gamma <- rnorm(n = nsites, mean = 2, sd = .2) # 10 site-level means (log(mean parasite load))
site.means.df <- data.frame(SiteCode = levels(ponddata$SiteCode), ExpMean = site.means.gamma)
ponddata2 <- merge(ponddata, site.means.df, by = "SiteCode")
ponddata2$Echinostoma <- rnbinom(n = nrow(ponddata2), mu = exp(ponddata2$ExpMean), size = 1)
stan.fit.sim1 <- stan_glmer(Echinostoma~ (1|SiteCode), data = ponddata2,  family =neg_binomial_2(link="log"))
rand.eff.sim1 <- coefficients(stan.fit.sim1)$SiteCode
{plot(rand.eff.sim1[,1]~site.means.gamma, xlab = "Actual Random Effects", ylab = "Estimated Random Effects", xlim = c(1.5,2.5), ylim = c(1.5, 2.5))
abline(0,1)} 
```

It's hard to repeat this varying things like mean, sd because the model takes so long to fit! oh well.
It also seems like the model underestimates variation in random effects--they are shrunk on the y axis...is this a partial pooling thing? A prior thing? 


Let's try this with  smaller variation between sites
```{r}
set.seed(1491)
site.means.gamma <- rnorm(n = nsites, mean = 2, sd = .05) # 10 site-level means (log(mean parasite load))
site.means.df <- data.frame(SiteCode = levels(ponddata$SiteCode), ExpMean = site.means.gamma)
ponddata2 <- merge(ponddata, site.means.df, by = "SiteCode")
ponddata2$Echinostoma <- rnbinom(n = nrow(ponddata2), mu = exp(ponddata2$ExpMean), size = 1)
stan.fit.sim1 <- stan_glmer(Echinostoma~ (1|SiteCode), data = ponddata2,  family =neg_binomial_2(link="log"))
rand.eff.sim1 <- coefficients(stan.fit.sim1)$SiteCode
{plot(rand.eff.sim1[,1]~site.means.gamma, xlab = "Actual Random Effects", ylab = "Estimated Random Effects", xlim = c(1.5, 2.5), ylim = c(1.5, 2.5))
abline(0,1)}
```



Try with a poisson instead of negative binomial
```{r}
set.seed(241)
site.means.gamma <- rnorm(n = nsites, mean = 2, sd = .5) # 10 site-level means (log(mean parasite load))
site.means.df <- data.frame(SiteCode = levels(ponddata$SiteCode), ExpMean = site.means.gamma)
ponddata2 <- merge(ponddata, site.means.df, by = "SiteCode")
ponddata2$Echinostoma <- rpois(n = nrow(ponddata2), lambda = exp(ponddata2$ExpMean))
stan.fit.sim1 <- stan_glmer(Echinostoma ~ (1|SiteCode), data = ponddata2,  family =poisson(link="log"))
rand.eff.sim1 <- coefficients(stan.fit.sim1)$SiteCode
{plot(rand.eff.sim1[,1]~site.means.gamma, xlab = "Actual Random Effects", ylab = "Estimated Random Effects")
abline(0,1)}
summary(stan.fit.sim1, digits = 3)
```

I think the dispersion parameter is leading to a lot of noise. The Poisson fit is better. There is an extra parameter to estimate with the negative binomial, so maybe that is the issue? Why is negative binomial so crappy?? 

The SD estimate around the intercept seems low (0.1)
The mean is right (2)


## SIMULATION 2: Random effects of site and sample
In this simulation we have a nested random effect for the collection visit (5 per site). The visit effects are drawn randomly from the mean for that site

```{r}
set.seed(523)
site.means.gamma <- rnorm(n = nsites, mean = 3, sd = .5) # 10 site-level means (log(mean parasite load))
visit.site.means <- data.frame(SiteCode = NA, visit = 1:5, visitmean = NA)
visit.site.means <- rep(list(visit.site.means), 10)
for(i in 1:nsites){
  visit.site.means[[i]][,1]<- levels(ponddata$SiteCode)[i]
  visit.site.means[[i]][,3] <- rnorm(n = 5, mean = site.means.gamma[i], sd = 0.2)

} # visit specific random effects (the site level re will change each time I run this code)
# since I didn't figure out the set seed thing within a loop
# don't want the visit effects to be the same per site

# get these in df form to combine with predictors
site.means.df <- data.frame(SiteCode = levels(ponddata$SiteCode), ExpMean = site.means.gamma)
visit.site.means <- do.call(rbind, visit.site.means)

# merge these into the dataframe
ponddata2 <- merge(ponddata, site.means.df, by = "SiteCode")
ponddata2 <- merge(ponddata2, visit.site.means, by = c("SiteCode", "visit") )

```

Simulate the expected number of parasites from the visit mean
```{r}
ponddata2$Echinostoma <- rnbinom(n = nrow(ponddata2), mu = exp(ponddata2$visitmean), size = 2)
```

Fit the model
```{r}
ponddata2$SiteCodevisit <- paste(ponddata2$SiteCode, "_", ponddata2$visit, sep = "")
stan.fit.sim2 <- stan_glmer(Echinostoma ~ (1|SiteCode) + (1|SiteCodevisit), data = ponddata2,  family =neg_binomial_2(link="log"))
```

Examine fit
```{r}
## site level
rand.eff.sim2.sc <- coefficients(stan.fit.sim2)$SiteCode
rand.eff.sim2.sv <- coefficients(stan.fit.sim2)$SiteCodevisit

{plot(rand.eff.sim2.sc[,1]~site.means.gamma, xlab = "Actual Random Effects", ylab = "Estimated Random Effects", main = "Site Level Random Effects")
abline(0,1)}  # looks good!


## visit level
visit.site.means$SiteCodevisit <- paste(visit.site.means$SiteCode,"_", visit.site.means$visit, sep = "")
rand.eff.sim2.sv$SiteCodevisit <- rownames(rand.eff.sim2.sv)
rand.eff.sim2.sv <- merge(rand.eff.sim2.sv, visit.site.means, by = "SiteCodevisit")
{plot(visitmean~`(Intercept)`, rand.eff.sim2.sv, xlab = "Actual Random Effects", ylab = "Estimated Random Effects")
abline(0,1)}  #looks decent or at least unbiased
```


## SIMULATION 3: Random effects of site and sample, and fixed effects of species and visit
This has a fixed effect for species (TATO have less parasites) and visit (parasite load goes up with time)

Set parameters
```{r}
nsites <- length(levels(ponddata$SiteCode)) # number of levels of the random effect (10)
site.means.gamma <- NA
set.seed(81)
site.means.gamma <- rnorm(n = nsites, mean = 2, sd = .2) # 10 site-level means (log(mean parasite load))
visit.site.means <- data.frame(SiteCode = NA, visit = 1:5, visitmean = NA)
visit.site.means <- rep(list(visit.site.means), 10)
visit.effect <- 0.2 # fixed effect for visit
species.effect <- -0.4 # fixed effect for species
for(i in 1:nsites){
  visit.site.means[[i]][,1]<- levels(ponddata$SiteCode)[i]
  visit.site.means[[i]][,3] <- rnorm(n = 5, mean = site.means.gamma[i], sd = 0.2)
  
} # visit specific random effects drawn from site means


# combine into a df with predictors
site.means.df <- data.frame(SiteCode = levels(ponddata$SiteCode), ExpMean = site.means.gamma)
visit.site.means <- do.call(rbind, visit.site.means)
# merge these into the dataframe
ponddata2 <- merge(ponddata, site.means.df, by = "SiteCode")
ponddata2 <- merge(ponddata2, visit.site.means, by = c("SiteCode", "visit") )
```

Simulate the expected number of parasites from the sample mean. The sample mean will become the intercept and I will add the fixed effects.
```{r}
# add fixed effect for visit and species (have to treat as numeric to multiply by species slope)
ponddata2$EchinostomaExp <- ponddata2$visitmean + ponddata2$visit*visit.effect + (as.numeric(ponddata2$SpeciesCode)-1)*species.effect
```

Generate observed parasite counts (observations) from the expected means, again, using some kind of overdispersion parameter
```{r}
ponddata2$EchinostomaObs<- rnbinom(n = nrow(ponddata2), mu = exp(ponddata2$EchinostomaExp), size = 2)
```

Fit the model
```{r}
ponddata2$SiteCodevisit <- paste(ponddata2$SiteCode, "_", ponddata2$visit, sep = "")
stan.fit.sim3 <- stan_glmer(EchinostomaObs ~ visit + SpeciesCode + (1|SiteCode) + (1|SiteCodevisit), data = ponddata2,  family =neg_binomial_2(link="log"))
```

Examine fit
```{r}
summary(stan.fit.sim3, digits = 4)
rand.eff.sim3.sc <- coefficients(stan.fit.sim3)$SiteCode
rand.eff.sim3.sv <- coefficients(stan.fit.sim3)$SiteCodevisit

## site level RE
{plot(rand.eff.sim3.sc[,1]~site.means.gamma, xlab = "Actual Random Effects", ylab = "Estimated Random Effects", xlim = c(1,2.5), ylim = c(1, 2.5), main = "Site level random effects")
abline(0,1)}  # looks bad! It's estimating a very low site-level variance! 

## visit level RE
visit.site.means$SiteCodevisit <- paste(visit.site.means$SiteCode,"_", visit.site.means$visit, sep = "")
rand.eff.sim3.sv$SiteCodevisit <- rownames(rand.eff.sim3.sv)
rand.eff.sim3.sv <- merge(rand.eff.sim3.sv, visit.site.means, by = "SiteCodevisit")
{plot(visitmean~`(Intercept)`, rand.eff.sim3.sv, xlab = "Actual Random Effects", ylab = "Estimated Random Effects", main = "Visit level random effects")
abline(0,1)} 


## fixed effects
stan.fit.sim3$coefficients[1:3]
# estimated visit effect is 0.186; compared to 0.2
# estimated species effect is -0.46; compared to -0.4
```


## Overall impressions
The model is pretty good at estimating fixed effects, but not so good at the random effects. Is this due to partial pooling? It's estimating a pretty low between site variance and the random effects per site are pretty small. Could be a coding error...?

