---
title: "McElreath_Ch4_Problems"
author: "Wynne Moss"
date: "September 17, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(rethinking)
```


## Easy problems
# 4E1) which line is the likelihood?
likelihood is the first line (data is drawn from the normal distribution centered around mu
# 4E2) how many parameters?
parameters: mu, sigma
# 4E3) Using the model definition above, write down the appropriate form of Bayes’ theorem thatincludes the proper likelihood and priors.
prob(mu, sigma |data) <- [prob(data | mu, sigma) x prob(mu) x prob (sigma)] /prob(data) over all values
# 4E4) which line is the linear model?
mu = alpha + beta * X

#4E5) how many parameters?
parameters: alpha, beta, sigma

## MEdium problems
#4M1) simulate observed heights from the prior
```{r}
# get random sigmas
sigmas <- runif(n = 1000, min = 0, max = 10)
# get random mean heights
mus <- rnorm (n = 1000, mean = 0, sd = 10)
# get random observed heights
ys <- rnorm(n = 1000, mean = mus,  sd = sigmas)
hist(ys)
dens(ys)
```

#4M2) Translate into an alist
```{r}
hlist <- alist(
  height ~ dnorm(mu, sigma), 
  mu ~ dnorm(0, 10),
  sigma ~ dunif(0, 10)
)
```

#4M3) translate alist into a mathematical formula
yi ~ Normal(mu, sigma)
mu = a + beta*xi
a ~ Normal(0, 50)
b ~ Uniform (0, 10)
sigma ~ Uniform(0, 50)

#4M4) formula for predicting height by hear
Height ~ Normal(mu, sigma)
mu ~ a + b*Year
a ~ Normal(0, 50)
b ~ Normal(0, 10)
sigma ~ Uniform(0, 50)

#4M5) adding more specific priors
I might use a uniform prior on beta that is  > 0
I might center my prior on sigma by 120 (but probably not since I don't thin you should use the same dataset to fit the piriors?)

#4M6) prior for sigma
I could use a uniform prior for sigma: dunif(0, 64)

## Hard probelms
4H1) Provide predicted heights using the kung model 
```{r}
# first fit the kung model
data(Howell1)
d <- Howell1
d2 <- d[ d$age >= 18 , ]
m4.3 <- sampost(
  alist(
    height ~ dnorm(mu, sigma),
    mu <- a + b*weight,
    a ~ dnorm(156, 100),
    b ~ dnorm(0, 10),
    sigma ~ dunif(0,50)
  ),
  data = d2
)
new.weights <- c(46.95, 43.72, 64.78, 32.59, 54.63)
n <- length(new.weights)
pred_hpdi_m <- matrix(NA,nrow = n,ncol=3)
colnames(pred_hpdi_m) <- c("low89","expected", "high89")
for(i in 1:n){
  mu <- m4.3$a + m4.3$b*new.weights[i] #the posterior sample of mu at weight w 
  newdat <- rnorm(n = length(mu), mu , sd = m4.3$sigma) # pull an observation from each value of sigma
  pred_hpdi_m[i, 1] <- HPDI(newdat,prob=0.89)[1] # hpdi of the sample 
  pred_hpdi_m[i, 2] <- mean(mu)
  pred_hpdi_m[i, 3] <- HPDI(newdat,prob=0.89)[2]
}
pred_hpdi_m
```

4H2) Fit a model using < 18 data
```{r}
d3 <- d[ d$age < 18 , ]
m4.5 <- sampost(
  alist(
    height ~ dnorm(mu, sigma),
    mu <- a + b*weight,
    a ~ dnorm(120, 100),
    b ~ dnorm(0, 10),
    sigma ~ dunif(0,50)
  ),
  data = d3
)
precis(m4.5)
```

(a) For every 10 unit increase in weight, children get 27.2 in taller

(b) Plot the data
```{r}
w <- seq (from = 0, to = 50, by = 1) # potential weights over which we want to plot best fit line
n <- length(w)
hdpi_m <- matrix(NA, nrow = n, ncol = 2) # matrix to store hdpi values
colnames(hdpi_m) <- c("low89", "high89")
for (i in 1:n){
  mu <- m4.5$a + m4.5$b*w[i] # the posterior sample of mu at weight i (for all posterior draws of a and b)
  hdpi_m[i, ] <- HPDI(mu, prob = 0.89)
}
{plot(height~weight, d3, col = "blue")
abline(a = mean(m4.5[, "a"]), b = mean(m4.5[,"b"]), col = "red")
lines(w, hdpi_m[, "low89"], col = "grey")
lines(w, hdpi_m[, "high89"], col = "grey")}
```


(c) The relationship looks curvilinear. I would probably change the assumption of them being linear and add a polynomial or log regression line.

4H2) Fitting a log-relationship
```{r}
m4.6 <- sampost(
  alist(
    height ~ dnorm(mu, sigma),
    mu <- a + b*log(weight),
    a ~ dnorm(178, 100),
    b ~ dnorm(0, 100),
    sigma ~ dunif(0,50)
  ),
  data = Howell1
)
precis(m4.6)
# plot the predicted mean and 97% HPDI
w <- seq (from = 1, to = 70, by = 1) # potential weights over which we want to plot best fit line
n <- length(w)
hdpi_m <- matrix(NA, nrow = n, ncol = 3) # matrix to store hdpi values
colnames(hdpi_m) <- c("low89", "high89", "mean")
for (i in 1:n){
  mu <- m4.6$a + m4.6$b*log(w[i]) # the posterior sample of mu at weight i (for all posterior draws of a and b)
  hdpi_m[i, 1:2 ] <- HPDI(mu, prob = 0.97)
  hdpi_m[i, 3] <- mean(mu)
}

{plot(height~weight, Howell1, col = "blue")
lines(w, hdpi_m[, "low89"], col = "grey")
lines(w, hdpi_m[, "high89"], col = "grey")
lines(w, hdpi_m[, "mean"], col="red")}
plot(height~weight, data = Howell1, col = col.alpha(rangi2, 0.4))
```


