---
title: "Simulation of parasite data for model checks"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rstan)
library(rstanarm)
library(tidyverse)
options(mc.cores = parallel::detectCores())
```

## Read in and clean up data
```{r}
ponddata <- read.csv("diss.data.2017.csv")
ponddata <- subset(ponddata, visit > 1) # take out first visit # fix this??
ponddata$visit <- ponddata$visit-1
ponddata <- ponddata[, c(2:6, 17,18)] # get just predictor variables
ponddata$visitScaledSVL <- rnorm(nrow(ponddata), 0, 2)# just make up this predictor 
```

## SIMULATION 1: Site-level random effect
This model predicts parasite richness using only a random effect for site.
Generate random effects for each of 10 sites from a normal distribution
```{r}
nsites <- length(levels(ponddata$SiteCode)) # number of levels of the random effect (10)
site.means.gamma <- NA
set.seed(12398) # makes sure the random draws are repeatable (?)

site.means.gamma <- rnorm(n = nsites, mean = 0, sd = .2) # 10 site-level means (log(mean parasite load))

# make a dataframe that we can merge with the predictor data frame
site.means.df <- data.frame(SiteCode = levels(ponddata$SiteCode), ExpMean = site.means.gamma)
# merge this with the existing predictors
ponddata2 <- merge(ponddata, site.means.df, by = "SiteCode")
```

Generate an observed parasite count from the expected site level mean. This is done using a poisson distribution.
```{r}
# for each row of the dataset, draw a random parasite richness count
# from the expected site level means
ponddata2$tot.para <- rpois(n = nrow(ponddata2), lambda = exp(ponddata2$ExpMean))
```

Fit a model; does it recapture the random effects?
```{r}
stan.fit.sim1 <- stan_glmer(tot.para~ (1|SiteCode), data = ponddata2,  family ="poisson")
summary(stan.fit.sim1)
rand.eff.sim1 <- coefficients(stan.fit.sim1)$SiteCode
rand.eff.sim1
site.means.gamma
```

Plot the estimates vs. real
```{r}
{plot(rand.eff.sim1[,1]~site.means.gamma, xlab = "Actual Random Effects", ylab = "Estimated Random Effects", xlim = c(-.5,.5), ylim = c(-0.5,.5))
abline(0,1)}
```


These look OK, now try again with diff set of random intercepts:
```{r}
set.seed(231)
site.means.gamma <- rnorm(n = nsites, mean = 0, sd = .2) # 10 site-level means (log(mean parasite richness))
site.means.df <- data.frame(SiteCode = levels(ponddata$SiteCode), ExpMean = site.means.gamma)
ponddata2 <- merge(ponddata, site.means.df, by = "SiteCode")
ponddata2$tot.para <- rpois(n = nrow(ponddata2), lambda = exp(ponddata2$ExpMean))
stan.fit.sim1 <- stan_glmer(tot.para~ (1|SiteCode), data = ponddata2,  family ="poisson")
rand.eff.sim1 <- coefficients(stan.fit.sim1)$SiteCode
{plot(rand.eff.sim1[,1]~site.means.gamma, xlab = "Actual Random Effects", ylab = "Estimated Random Effects", xlim = c(-.5,.5), ylim = c(-.5, .5))
abline(0,1)} 
```

It's hard to repeat this varying things like mean, sd because the model takes so long to fit! oh well.
It also seems like the model underestimates variation in random effects--they are shrunk on the y axis...is this a partial pooling thing? A prior thing? 


Let's try this with  smaller variation between sites
```{r}
set.seed(1491)
site.means.gamma <- rnorm(n = nsites, mean = 0, sd = .05) # 10 site-level means (log(mean parasite load))
site.means.df <- data.frame(SiteCode = levels(ponddata$SiteCode), ExpMean = site.means.gamma)
ponddata2 <- merge(ponddata, site.means.df, by = "SiteCode")
ponddata2$tot.para <- rpois(n = nrow(ponddata2), lambda = exp(ponddata2$ExpMean))
stan.fit.sim1 <- stan_glmer(tot.para~ (1|SiteCode), data = ponddata2,  family ="poisson")
rand.eff.sim1 <- coefficients(stan.fit.sim1)$SiteCode
{plot(rand.eff.sim1[,1]~site.means.gamma, xlab = "Actual Random Effects", ylab = "Estimated Random Effects", xlim = c(-.5,.5), ylim = c(-.5,.5))
abline(0,1)}
```


## SIMULATION 2: Random effects of site and sample
In this simulation we have a nested random effect for the collection visit (5 per site). The visit effects are drawn randomly from the mean for that site

```{r}
set.seed(523)
site.means.gamma <- rnorm(n = nsites, mean = 0, sd = .5) # 10 site-level means (log(mean parasite load))
visit.site.means <- data.frame(SiteCode = NA, visit = 1:5, visitmean = NA)
visit.site.means <- rep(list(visit.site.means), 10)
for(i in 1:nsites){
  visit.site.means[[i]][,1]<- levels(ponddata$SiteCode)[i]
  visit.site.means[[i]][,3] <- rnorm(n = 5, mean = site.means.gamma[i], sd = 0.2)

} # visit specific random effects (the site level re will change each time I run this code)
# since I didn't figure out the set seed thing within a loop
# don't want the visit effects to be the same per site

# get these in df form to combine with predictors
site.means.df <- data.frame(SiteCode = levels(ponddata$SiteCode), ExpMean = site.means.gamma)
visit.site.means <- do.call(rbind, visit.site.means)

# merge these into the dataframe
ponddata2 <- merge(ponddata, site.means.df, by = "SiteCode")
ponddata2 <- merge(ponddata2, visit.site.means, by = c("SiteCode", "visit") )

```

Simulate the expected number of parasites from the visit mean
```{r}
ponddata2$tot.para <- rpois(n = nrow(ponddata2), lambda = exp(ponddata2$visitmean))
```

Fit the model
```{r, message = FALSE}
ponddata2$SiteCodevisit <- paste(ponddata2$SiteCode, "_", ponddata2$visit, sep = "")
stan.fit.sim2 <- stan_glmer(tot.para ~ (1|SiteCode) + (1|SiteCodevisit), data = ponddata2,  family = "poisson")
```

Examine fit
```{r}
## site level
rand.eff.sim2.sc <- coefficients(stan.fit.sim2)$SiteCode
rand.eff.sim2.sv <- coefficients(stan.fit.sim2)$SiteCodevisit

{plot(rand.eff.sim2.sc[,1]~site.means.gamma, xlab = "Actual Random Effects", ylab = "Estimated Random Effects", main = "Site Level Random Effects", xlim = c(-1.5,1.5), ylim = c(-1.5,1.5))
abline(0,1)}  # looks good!


## visit level
visit.site.means$SiteCodevisit <- paste(visit.site.means$SiteCode,"_", visit.site.means$visit, sep = "")
rand.eff.sim2.sv$SiteCodevisit <- rownames(rand.eff.sim2.sv)
rand.eff.sim2.sv <- merge(rand.eff.sim2.sv, visit.site.means, by = "SiteCodevisit")
{plot(visitmean~`(Intercept)`, rand.eff.sim2.sv, xlab = "Actual Random Effects", ylab = "Estimated Random Effects", xlim = c(-1,1), ylim = c(-1,1))
abline(0,1)}  #looks less great
```

These are pretty big random effects and the deviation is worrisome...


## SIMULATION 3: Random effects of site and sample, and fixed effects of species and visit
richness ~ visitscaledSVL + visit + visit^2 + speciesCode + speciesCode*visit
Set parameters
```{r}
nsites <- length(levels(ponddata$SiteCode)) # number of levels of the random effect (10)
site.means.gamma <- NA
set.seed(122)
site.means.gamma <- rnorm(n = nsites, mean = 0, sd = .5) # 10 site-level means (log(mean parasite load))
visit.site.means <- data.frame(SiteCode = NA, visit = 1:5, visitmean = NA)
visit.site.means <- rep(list(visit.site.means), 10)

## set fixed parameters
visit.effect <- 0.85 # fixed effect for visit
visit.effect.sq <- -0.1
species.effect <- -0.4 # fixed effect for species
visit.species.effect <- -1 # fixed effect for species
svl.effect <- 0.3
Effects <- data.frame(FixedEffect = c("Visit", "Species", "Visit^2", "SVL", "Visit*Species"), 
             TrueValue =  c(visit.effect, species.effect, visit.effect.sq, svl.effect, visit.species.effect))

for(i in 1:nsites){
  visit.site.means[[i]][,1]<- levels(ponddata$SiteCode)[i]
  for(j in 1:5){
    Exp.mean <- site.means.gamma[i] + j*visit.effect + j^2*visit.effect.sq
    visit.site.means[[i]][j,3] <- rnorm(n=1, mean = Exp.mean, sd = .1)
  }
} # visit specific random effects drawn from expected mean. Putting the visit effect here. 


# combine into a df with predictors
site.means.df <- data.frame(SiteCode = levels(ponddata$SiteCode), ExpMean = site.means.gamma)
visit.site.means <- do.call(rbind, visit.site.means)
# merge these into the dataframe
ponddata2 <- merge(ponddata, site.means.df, by = "SiteCode")
ponddata2 <- merge(ponddata2, visit.site.means, by = c("SiteCode", "visit") )
```

Simulate the expected number of parasites from the sample mean. The sample mean will become the intercept and I will add the fixed effects.
```{r}
# add fixed effect for visit and species (have to treat as numeric to multiply by species slope)
ponddata2$SpeciesCode <- as.numeric(ponddata2$SpeciesCode)-1

ponddata2$ParasiteRichExp <-  ponddata2$visitmean +
                              ponddata2$visitScaledSVL*svl.effect + 
                              ponddata2$SpeciesCode*species.effect +   
                              ponddata2$SpeciesCode*ponddata2$visit*visit.species.effect

```

Generate observed parasite counts (observations) from the expected means, again, using some kind of overdispersion parameter
```{r}
ponddata2$ParasiteRichObs<- rpois(n = nrow(ponddata2), lambda = exp(ponddata2$ParasiteRichExp))
```

Fit the model
```{r, message = FALSE, warning = FALSE}
ponddata2$SiteCodevisit <- paste(ponddata2$SiteCode, "_", ponddata2$visit, sep = "")
stan.fit.sim3 <- stan_glmer(ParasiteRichObs ~ visit*SpeciesCode + I(visit^2) + visitScaledSVL +  (1|SiteCode) + (1|SiteCodevisit), data = ponddata2,  family = "poisson")
```

Examine fit
```{r}
summary(stan.fit.sim3, digits = 4)
Effects$EstimatedValue <- as.vector(stan.fit.sim3$coefficients[2:6])
Effects$EstimatedValueSE <- stan.fit.sim3$ses[2:6]

rand.eff.sim3.sc <- coefficients(stan.fit.sim3)$SiteCode
rand.eff.sim3.sv <- coefficients(stan.fit.sim3)$SiteCodevisit

## site level RE
{plot(rand.eff.sim3.sc[,1]~site.means.gamma, xlab = "Actual Random Effects", ylab = "Estimated Random Effects", main = "Site level random effects")
abline(0,1)}  # looks pretty good! 


ggplot(Effects)+
  geom_hline(yintercept = 0, linetype = "dashed", col = "black")+
  geom_point(aes(x=FixedEffect, y = EstimatedValue, colour = "EstimatedValue"), shape = 16, size = 2)+
  geom_errorbar(aes(x=FixedEffect,ymin = EstimatedValue - 2*EstimatedValueSE, ymax = EstimatedValue + 2*EstimatedValueSE, colour = "EstimatedValue"), width = .1)+ 
  geom_point(aes(x=FixedEffect, y = TrueValue, colour = "TrueValue"), shape = 95, size = 10)+
  scale_colour_manual(name = "", values = c(EstimatedValue="dodgerblue", TrueValue = "red"), labels = c("Estimated Value", "True Value"))+
  xlab("Parameter")+
  ylab("Coefficient Value")+
  theme(legend.position = "bottom")
```



## Overall impressions
The model is pretty good at estimating fixed effects and the site-level random effects. I'm not really sure how to plot the site-visit random effects since they are a combination of fixed predictors and random effects. However, I think this output confirms that the model is doing a pretty decent job at estimation. I haven't played around with standard deviations much to see how this impacts estimation, since the model takes so long to fit. 
